Bayesian hidden Markov model analysis of single-molecule force spectroscopy:\\ Characterizing kinetics under measurement uncertainty

### 2015-05-17 manuscripts/bayesian-hmm-method/bhmm.tex
Updated synthetic three-state model plot- and table-generation scripts; updates to manuscript.

@@

\author{Frank No\'{e}}
 \email{frank.noe@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

@@

Single-molecule force spectroscopy has proven to be a powerful tool for studying the kinetic behavior of biomolecules.
Through application of an external force, conformational states with small or transient populations can be stabilized, allowing them to be characterized and the statistics of individual trajectories studied to provide insight into biomolecular folding and function.
Because the observed quantity (force or extension) is not necessarily an ideal reaction coordinate, individual observations cannot be uniquely associated with kinetically distinct conformations.
While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and permits uncertainty to be further reduced through the imposition of physical priors (such as the requirement that dynamics obey detailed balance).}
We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}

@@

Recent advances in biophysical measurement have led to an unprecedented ability to monitor the dynamics of single biological macromolecules, such as proteins and nucleic acids~\cite{ritort:j-phys:2006:single-molecule-review}.
%As a new approach to probing the behavior of biological macromolecules, these experiments promise to change the way we study folding, dynamics, catalysis, association, transcription, translation, and motility, providing otherwise-inaccessible information about microscopic kinetics, energetics, mechanism, and the stochastic heterogeneity inherent in these processes.
Advances in instrumentation for optical force spectroscopy in particular have produced instruments of extraordinary stability, precision, and temporal resolution~\cite{block:rev-sci-instrum:2004:optical-trapping,bustamante:annu-rev-biochem:2008:optical-tweezers} that have already demonstrated great utility in the study of biomolecules in the presence of externally perturbative forces~\cite{bustamante:nature:2003:dna-mechanics,block:curr-opin-chem-biol:2008:rna,bustamante:cell:2011:central-dogma}. 
Under external force, it becomes possible to stabilize and characterize short-lived conformational states, such as protein folding and unfolding intermediates~\cite{clarke:jmb:2002:titin-afm,cecconi-shank-bustamante-marqusee:science:2005:rnase-h,rief:pnas:2010:pmf}.

In a typical single-molecule optical trapping experiment, a protein or nucleic acid is tethered to two polystyrene beads by dsDNA handles that {\color{red}serve as spacers to} prevent the molecule under study from interacting with the beads {({\color{red} as in Figure~\ref{figure:trap-configuration})}.
The handle-biomolecule-handle assembly---referred to as a fiber---is associated with the beads through tight noncovalent interactions, with one bead held in an optical trap and the other either suctioned to a micropipette or held in a second optical trap.
During an experiment, the position of the bead within the laser trap is monitored, and either the relative displacement from the trap center or the total force on the bead is recorded, resulting in a timeseries such as the one depicted in Fig.~\ref{figure:p5ab-observed-trajectory}.
{\color{red}The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system {\color{red}transiently} out of equilibrium; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber.}
Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review,elms:biophys-j:2012:force-feedback}
{\color{magenta}FN: add additional citations for both modes / analyses}.

Often, the dynamics observed in these experiments appears to be dominated by stochastic transitions between two or more strongly metastable conformational states~\cite{schuette-huisinga:2002:biomolecular-conformations-as-metastable-states,chodera-singhal:jcp:2007:automatic-state-decomposition}---regions of conformation space in which the system remains for long times before making a transition to another conformational state.
{\color{red}The observed force trace in Fig.~\ref{figure:p5ab-observed-trajectory}, for example, demonstrates apparent hopping behavior between two or more such states.}
{\color{red}Transitions among strongly metastable states such as these} are generally well-described by first-order kinetics~\cite{noe:pnas:2009:ww-domain}.
While visual inspection of the dynamics may suggest the clear presence of multiple metastable states, quantitative characterization of these states is often difficult.
First, the observed force or extension is unlikely to correspond to a true reaction coordinate easily able to separate all metastable states~\cite{best-paci-hummer-dudko:jpcb:2008:pulling-coordinate,rief-paci:pre:2010:multidimensional-landscape,thirumalai:2011:prl:pfold,chodera-pande:arxiv:2011:single-molecule-pfold}, and second, measurement noise may further broaden the force or extension signatures of individual states, increasing their overlap.
{\color{red} As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in Fig.~\ref{figure:p5ab-observed-trajectory}, right.}
Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory,PrinzChoderaNoe_PRX14_RateTheory} (see \emph{Supplementary Material: Comparison with threshold model}).

@@

\begin{figure}[tbp]
\noindent
\resizebox{\columnwidth}{!}{\includegraphics{optical-trap-configuration-threestate-trimmed}}
\caption{{\bf Single-molecule optical trapping configuration.}
The biomolecule of interest---here, %the p5ab 
{\color{red} an} RNA hairpin---is tethered to two polystyrene beads by dsDNA handles.
The fluctuating force on one bead held in an optical trap is monitored, while the other bead is held suctioned to a micropipette tip.
Conformational transitions of the hairpin---such as transitions among the three kinetically metastable states illustrated here---are observed indirectly through motion of the bead in the trap.
\label{figure:trap-configuration}}
\end{figure}

Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data {\color{red} (used to color the points in Fig.~\ref{figure:p5ab-observed-trajectory})} can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
Unfortunately, this approach has a number of significant drawbacks.
%Due to technical limitations, experiments often suffer from limited statistics---the events of interest (transitions between states or visits to rare states) may occur only a few times during the course of the measurement, and data for additional fibers is time-consuming to collect.
%As a result, while the MLE yields the most likely set of model parameters, there may be 
{\color{magenta}The parameters estimated by the MLE might be subject to enormous uncertainty. 
Methods for estimating the standard error or confidence intervals from hidden Markov models~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error} make a number of approximations that can lead to a significant underestimation of the error.}
%, these schemes can be prohibitively costly for long traces, and may still significantly underestimate the statistical error for short traces due to the normally-distributed error approximation inherent in the approach.
%{\color{blue}[JDC: Should we say something about bootstrap here since we will compare to it later?]}
%The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
Worse yet, the standard algorithms employed to find the MLE may not even find the true maximum likelihood solution, instead converging to a local maximum in likelihood that is far from optimal~\cite{merialdo:ieee-trans-speech-audio-proc:1993:baum-welch-localit}.

@@

{\color{red}To demonstrate the utility of the Bayesian HMM in the analysis of real force spectroscopy data, we illustrate its application to the analysis of an RNA hairpin exhibiting apparent three-state behavior (Figure~\ref{figure:p5ab-observed-trajectory}) and a protein system that exhibits two-state behavior with highly overlapping force signatures (Figure~\ref{figure:apomyoglobin-observed-trajectory}) collected in a passive (equilibrium) optical trap.
In addition to maximum likelihood estimates of equilibrium state probabilities, state-to-state transition rates, and the force distributions characterizing each state, the Bayesian extension provides estimates of confidence intervals for individual parameters (Tables~\ref{table:p5ab-bhmm-confidence-intervals} and \ref{table:apomyoglobin-bhmm-confidence-intervals}) as well as the full joint distribution of parameters.

%In the following sections, we briefly describe the theory behind maximum-likelihood hidden Markov models of force spectroscopy data (\emph{Theory}), the validation of our methodology on synthetic data for which the computed confidence intervals can be checked for correctness (\emph{Validation}), and our specific applications to real force spectroscopy data (\emph{Applications}).
%We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/choderalab/bhmm}].
}

@@

We presume the system under study has $M$ kinetically distinct states, in the sense that the system generally remains in a given state for several observation intervals $\Delta t$, but these states may not necessarily represent highly populated states of the system at equilibrium.
We treat these conformational states as the hidden states of the model, because we cannot directly observe the identity of the metastable state in which the system resides.
The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on {\color{red} the hidden states $\bfm{S} \equiv \{s_t\}$,} parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
\begin{eqnarray}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t \mid s^{}_{t-1}, \bfm{T} &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}} &\sim& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t &\sim& \varphi(\cdot \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
\Pr(s^{}_0) &=& \rho^{}_{s^{}_0} \nonumber \\
\Pr(s^{}_t \mid s^{}_{t-1}, \bfm{T}) &=& T_{s^{}_{t-1} s^{}_{t}} , \:\: t \ge 1 \nonumber \\
\Pr(o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}}) &=& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process} .
\end{eqnarray}
In diagrammatic form, the observed state data $\{o_t\}$ and corresponding hidden state history $\{s_t\}$ can be represented 
\begin{eqnarray}
\begin{array}{llllllllll}
 \stackrel{\bfv{\rho}}{\longrightarrow} & s^{}_{0} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{1} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{2} & \stackrel{\bfm{T}}{\longrightarrow} & \cdots & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{L^{}} \\
 & \downarrow \varphi &  & \downarrow \varphi & & \downarrow \varphi & & & & \downarrow \varphi \\
 & o^{}_{0} & & o^{}_{1} & & o^{}_{2} &  & & & o^{}_{L^{}} \\
\end{array}
\end{eqnarray} 
The initial state distribution $\bfv{\rho}^{}$  reflects our knowledge of the initial conditions of the experiment that collected data $\bfm{{\color{red} O}}^{}$.
In the case that the experiment was prepared in equilibrium, $\bfv{\rho}$ corresponds to the equilibrium distribution $\bfv{\pi}$ of the model transition matrix $\bfm{T}$; if the experiment was prepared out of equilibrium, $\bfv{\rho}$ may be chosen to reflect some other prior distribution (e.g.~the uniform prior).

@@

\begin{figure*}[tbp]
\noindent
\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
\caption{{\bf Synthetic force trajectory and inferred state assignments in MLHMM.}
Observed samples are colored by their hidden state assignments.
Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
The gray histogram on the right side shows the total observed probability of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
\label{figure:model-stateassignments}}
\end{figure*}

>>>>

\author{Frank No\'{e}}
 \thanks{Corresponding author}
 \email{frank.noe@fu-berlin.de}
 \affiliation{DFG Research Center Matheon, FU Berlin, Arnimallee 6, 14195 Berlin, Germany}

@@

@@

Single-molecule force spectroscopy has proven to be a powerful tool for studying the kinetic behavior of biomolecules.
Through application of an external force, conformational states with small or transient populations can be stabilized, allowing them to be characterized and the statistics of individual trajectories studied to provide insight into biomolecular folding and function.
Because the observed quantity (force or extension) is not necessarily an ideal reaction coordinate, individual observations cannot be uniquely associated with kinetically distinct conformations.
While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and permits uncertainty to be further reduced through the imposition of physical priors (such as the requirement that dynamics obey detailed balance).
We illustrate the utility of this approach in characterizing the observed multistate kinetics of several molecules in a stationary optical trap: a simple protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state, an RNA hairpin that exhibits three-state behavior, and a complex protein system (RNase H) that exhibits complex multistate behavior.

@@

Recent advances in biophysical measurement have led to an unprecedented ability to monitor the dynamics of single biological macromolecules, such as proteins and nucleic acids~\cite{ritort:j-phys:2006:single-molecule-review}.
Advances in instrumentation for optical force spectroscopy in particular have produced instruments of extraordinary stability, precision, and temporal resolution~\cite{block:rev-sci-instrum:2004:optical-trapping,bustamante:annu-rev-biochem:2008:optical-tweezers} that have already demonstrated great utility in the study of biomolecules in the presence of externally perturbative forces~\cite{bustamante:nature:2003:dna-mechanics,block:curr-opin-chem-biol:2008:rna,bustamante:cell:2011:central-dogma}. 
Under external force, it becomes possible to stabilize and characterize short-lived conformational states, such as protein folding and unfolding intermediates~\cite{clarke:jmb:2002:titin-afm,cecconi-shank-bustamante-marqusee:science:2005:rnase-h,rief:pnas:2010:pmf}.

In a typical single-molecule optical trapping experiment, a protein or nucleic acid is tethered to two polystyrene beads by dsDNA handles that serve as spacers to prevent the molecule under study from interacting with the beads (as in Figure~\ref{figure:trap-configuration}).
The handle-biomolecule-handle assembly---referred to as a fiber---is associated with the beads through tight noncovalent interactions, with one bead held in an optical trap and the other either suctioned to a micropipette or held in a second optical trap.
During an experiment, the position of the bead within the laser trap is monitored, and either the relative displacement from the trap center or the total force on the bead is recorded, resulting in a timeseries such as the one depicted in Fig.~\ref{figure:p5ab-observed-trajectory}.
The instrument can generally be operated in several modes: a force ramp mode, in which the trap is translated rapidly enough to potentially carry the system transiently out of equilibrium~\cite{force-ramp}; a passive mode, in which the trap is held fixed and the equilibrium fluctuations of the system are observed~\cite{passive-mode}; and a constant force-feedback mode, in which the trap is continually repositioned to maintain a specified average constant force on the fiber~\cite{force-feedback}.
Here, we concern ourselves with passive-mode force probe experiments, though both nonequilibrium experiments and constant force-feedback experiments and their analysis remain an exciting topic of active research~\cite{ritort:adv-chem-phys:2008:nonequilibrium-review,elms:biophys-j:2012:force-feedback,woodside-nonequilibrium}.
{\color{red}[FN: Add additional citations for these modes and analyses.]}

Often, the dynamics observed in these experiments appears to be dominated by stochastic transitions between two or more strongly metastable conformational states~\cite{schuette-huisinga:2002:biomolecular-conformations-as-metastable-states,chodera-singhal:jcp:2007:automatic-state-decomposition}---regions of conformation space in which the system remains for long times before making a transition to another conformational state.
The observed force trace in Fig.~\ref{figure:p5ab-observed-trajectory}, for example, demonstrates apparent hopping behavior between two or more such states.
Transitions among strongly metastable states such as these are generally well-described by first-order kinetics~\cite{noe:pnas:2009:ww-domain}.
While visual inspection of the dynamics may suggest the clear presence of multiple metastable states, quantitative characterization of these states is often difficult.
First, the observed force or extension is unlikely to correspond to a true reaction coordinate easily able to separate all metastable states~\cite{best-paci-hummer-dudko:jpcb:2008:pulling-coordinate,rief-paci:pre:2010:multidimensional-landscape,thirumalai:2011:prl:pfold,chodera-pande:arxiv:2011:single-molecule-pfold}, and second, measurement noise may further broaden the force or extension signatures of individual states, increasing their overlap.
As a result, a histogram of the observed temporal trace can have strongly overlapping state distributions, as in the right panel of Fig.~\ref{figure:p5ab-observed-trajectory}.
Attempting to separate these states by simply dividing the observed force or extension range into regions, following current practice~\cite{woodside:science:2006:dna-hairpin-optical-trap,ritort:biophys-j:2011:short-handles}, can often lead to a high degree of state mis-assignment that results in the estimated rate constants and state distributions containing a significant amount of error~\cite{chodera:rate-theory,PrinzChoderaNoe_PRX14_RateTheory} (see \emph{Supplementary Material: Comparison with threshold model}).

@@

\begin{figure}[tbp]
\noindent
\resizebox{\columnwidth}{!}{\includegraphics{optical-trap-configuration-threestate-trimmed}}
\caption{{\bf Single-molecule optical trapping configuration.}
The biomolecule of interest---here, an RNA hairpin---is tethered to two polystyrene beads by dsDNA handles.
The fluctuating force on one bead held in an optical trap is monitored, while the other bead is held suctioned to a micropipette tip.
Conformational transitions of the hairpin---such as transitions among the three kinetically metastable states illustrated here---are observed indirectly through motion of the bead in the trap.
\label{figure:trap-configuration}}
\end{figure}

Given observed timeseries data for forces or extensions, the maximum likelihood estimate (MLE) of the model parameters (transition rates and state force or extension distributions) and sequence of hidden states corresponding to the observed data (used to color the points by state assignment in Fig.~\ref{figure:p5ab-observed-trajectory}) can be determined by standard methods~\cite{baum:1970:ann-math-statist:baum-welch,viterbi:1967:IEEE-trans-info-theory:viterbi-algorithm}, as demonstrated in recent work~\cite{kruithof-vannoort:biophys-j:2009:hmm-force}.
Unfortunately, this approach has a number of significant drawbacks.
The parameters estimated by the MLE might be subject to enormous uncertainty, especially when the dataset contains a small number of transitions between states or visits to rare states.
Methods for estimating the standard error or confidence intervals from hidden Markov models~\cite{aittokallio-uusipaikka:technical-report:2000:hmm-standard-error} make a number of approximations that can lead to a significant underestimation of the error.
%The high cost (both in terms of instrument and experimenter time) of collecting additional data also means that it is not a simple task to judge how much data need be collected to test a particular hypothesis in a statistically meaningful way.
Worse yet, the standard algorithms employed to find the MLE may not even find the true maximum likelihood solution, instead converging to a local maximum in likelihood that is far from optimal~\cite{merialdo:ieee-trans-speech-audio-proc:1993:baum-welch-localit}.

@@

To demonstrate the utility of the Bayesian HMM in the analysis of real force spectroscopy data, we illustrate its application to the analysis of an RNA hairpin exhibiting apparent three-state behavior (Figure~\ref{figure:p5ab-observed-trajectory}) and a protein system that exhibits two-state behavior with highly overlapping force signatures (Figure~\ref{figure:apomyoglobin-observed-trajectory}) collected in a passive (equilibrium) optical trap.
In addition to maximum likelihood estimates of equilibrium state probabilities, state-to-state transition rates, and the force distributions characterizing each state, the Bayesian extension provides estimates of confidence intervals for individual parameters (Tables~\ref{table:p5ab-bhmm-confidence-intervals} and \ref{table:apomyoglobin-bhmm-confidence-intervals}) as well as the full joint distribution of parameters.

%In the following sections, we briefly describe the theory behind maximum-likelihood hidden Markov models of force spectroscopy data (\emph{Theory}), the validation of our methodology on synthetic data for which the computed confidence intervals can be checked for correctness (\emph{Validation}), and our specific applications to real force spectroscopy data (\emph{Applications}).
%We follow with a discussion of future extensions of this approach, and include a complete description of the computational implementation in \emph{Algorithms}.
A Matlab implementation of the approach described here is also made freely available online [\url{http://github.com/bhmm/bhmm}].

@@

We presume the system under study has $M$ kinetically distinct states, in the sense that the system generally remains in a given state for several observation intervals $\Delta t$, but these states may not necessarily represent highly populated states of the system at equilibrium.
We treat these conformational states as the hidden states of the model, because we cannot directly observe the identity of the metastable state in which the system resides.
The hidden Markov model presumes the observed data $\bfm{O} \equiv \{o^{}_t\}$ was generated according to the following model dependent on the sequence of hidden states $\bfm{S} \equiv \{s_t\}$, parameters $\bfm{\Theta} \equiv \{\bfm{T}, \bfm{E}\}$, where $\bfm{T}$ is an $M \times M$ row-stochastic transition matrix and $\bfm{E}$ a set of emission parameters governing the observable (force or extension) distributions for each of the $M$ hidden states, prior information about the initial state distribution $\bfv{\rho}^{}$, {\color{red} and emission probability distribution $\varphi(o | \bfm{e})$},
\begin{eqnarray}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t \mid s^{}_{t-1}, \bfm{T} &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}} &\sim& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
%s^{}_0 &\sim& \bfv{\rho} \nonumber \\
%s^{}_t &\sim& T_{s^{}_{t-1} \cdot} \nonumber \\
%o^{}_t &\sim& \varphi(\cdot \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process}
\Pr(s^{}_0) &=& \rho^{}_{s^{}_0} \nonumber \\
\Pr(s^{}_t \mid s^{}_{t-1}, \bfm{T}) &=& T_{s^{}_{t-1} s^{}_{t}} , \:\: t \ge 1 \nonumber \\
\Pr(o^{}_t \mid s^{}_t, \bfm{e}_{s^{}_{t}}) &=& \varphi(o^{}_{t} \mid \bfm{e}_{s^{}_{t}}) \label{equation:hmm-process} .
\end{eqnarray}
In diagrammatic form, the observed state data $\{o_t\}$ and corresponding hidden state history $\{s_t\}$ can be represented 
\begin{eqnarray}
\begin{array}{llllllllll}
 \stackrel{\bfv{\rho}}{\longrightarrow} & s^{}_{0} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{1} & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{2} & \stackrel{\bfm{T}}{\longrightarrow} & \cdots & \stackrel{\bfm{T}}{\longrightarrow} & s^{}_{L^{}} \\
 & \downarrow \varphi &  & \downarrow \varphi & & \downarrow \varphi & & & & \downarrow \varphi \\
 & o^{}_{0} & & o^{}_{1} & & o^{}_{2} &  & & & o^{}_{L^{}} \\
\end{array}
\end{eqnarray} 
The initial state distribution $\bfv{\rho}^{}$  reflects our knowledge of the initial conditions of the experiment that collected data $\bfm{{\color{red} O}}^{}$.
In the case that the experiment was prepared in equilibrium, $\bfv{\rho}$ corresponds to the equilibrium distribution $\bfv{\pi}$ of the model transition matrix $\bfm{T}$; if the experiment was prepared out of equilibrium, $\bfv{\rho}$ may be chosen to reflect some other prior distribution (e.g.~the uniform prior).

@@

\begin{figure*}[tbp]
\noindent
%\resizebox{\textwidth}{!}{\includegraphics{model-mlhmm-stateassignments}}
\resizebox{\textwidth}{!}{\includegraphics{synthetic-three-state-model-mlhmm}}
\caption{{\bf Synthetic force trajectory and inferred state assignments in MLHMM.}
Observed samples are colored by their hidden state assignments.
Dark horizontal lines terminating in triangles to the right denote state means, while lightly colored bands indicate one standard deviation on either side of the state mean.
The histograms on the right side shows the state-resolved probabilities of samples, while the colored peaks show the weighted Gaussian output contribution from each state, and the black outline the weighted sum of the Gaussian output contributions from the HMM states.
\label{figure:model-stateassignments}}
\end{figure*}

\begin{table*}
\caption{{\bf Estimated mean model parameters and confidence intervals for synthetic timeseries data}}
\label{table:synthetic-confidence-intervals}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lccccc}
\hline
&  &  & \multicolumn{3}{c}{\bf Estimated Model Parameters}  \\ \cline{4-6}
\multicolumn{2}{l}{\bf Property} & \bf True Value & \bf 1 000 observations & \bf 10 000 observations & \bf 100 000 observations\\ \hline
		Equilibrium probability 		& $\pi_{1}$ & $0.333$ & $0.338_{\:0.130}^{\:0.543}$  & $0.309_{\:0.251}^{\:0.374}$  & $0.330_{\:0.299}^{\:0.363}$  \\
		& $\pi_{2}$ & $0.333$ & $0.351_{\:0.082}^{\:0.735}$  & $0.335_{\:0.246}^{\:0.432}$  & $0.329_{\:0.309}^{\:0.350}$  \\
		& $\pi_{3}$ & $0.333$ & $0.312_{\:0.118}^{\:0.510}$  & $0.356_{\:0.299}^{\:0.411}$  & $0.341_{\:0.322}^{\:0.357}$  \\
		\hline
		Transition probability ($\Delta t = $1.0 ms) 		& $T_{11}$ & $0.989$ & $0.952_{\:0.926}^{\:0.972}$ & $0.947_{\:0.939}^{\:0.955}$ & $0.988_{\:0.987}^{\:0.989}$ \\
		& $T_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
		& $T_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
		& $T_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
		& $T_{22}$ & $0.940$ & $0.990_{\:0.975}^{\:0.998}$ & $0.988_{\:0.983}^{\:0.991}$ & $0.949_{\:0.946}^{\:0.951}$ \\
		& $T_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
		& $T_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
		& $T_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
		& $T_{33}$ & $0.949$ & $0.949_{\:0.922}^{\:0.970}$ & $0.945_{\:0.937}^{\:0.953}$ & $0.941_{\:0.938}^{\:0.943}$ \\
		\hline
		\hline
		Transition rate (ms$^{-1}$) 		& $k_{12}$ & $0.010$ & $0.005_{\:0.001}^{\:0.015}$ & $0.001_{\:0.001}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
		& $k_{13}$ & $0.001$ & $0.043_{\:0.024}^{\:0.067}$ & $0.051_{\:0.044}^{\:0.060}$ & $0.011_{\:0.010}^{\:0.012}$ \\
		& $k_{21}$ & $0.010$ & $0.005_{\:0.001}^{\:0.016}$ & $0.001_{\:0.000}^{\:0.003}$ & $0.001_{\:0.001}^{\:0.001}$ \\
		& $k_{23}$ & $0.050$ & $0.005_{\:0.001}^{\:0.015}$ & $0.011_{\:0.008}^{\:0.015}$ & $0.050_{\:0.048}^{\:0.053}$ \\
		& $k_{31}$ & $0.001$ & $0.046_{\:0.028}^{\:0.072}$ & $0.044_{\:0.038}^{\:0.052}$ & $0.011_{\:0.009}^{\:0.012}$ \\
		& $k_{32}$ & $0.050$ & $0.005_{\:0.001}^{\:0.016}$ & $0.010_{\:0.007}^{\:0.014}$ & $0.049_{\:0.046}^{\:0.051}$ \\
		\hline
		State mean lifetime (ms) 		& $t_{1}$ & $90.408$ & $21.797_{\:13.184}^{\:36.162}$ & $18.645_{\:15.899}^{\:21.524}$ & $83.405_{\:75.423}^{\:92.188}$ \\
		& $t_{2}$ & $16.162$ & $148.113_{\:43.004}^{\:525.609}$ & $82.451_{\:59.320}^{\:115.043}$ & $18.954_{\:18.001}^{\:19.919}$ \\
		& $t_{3}$ & $19.103$ & $20.138_{\:12.475}^{\:33.395}$ & $17.902_{\:15.417}^{\:20.637}$ & $16.395_{\:15.692}^{\:17.177}$ \\
		\hline
		Relaxation time (ms) 		& $\tau_{1}$ & $62.700$ & $82.300_{\:32.985}^{\:200.069}$ & $56.416_{\:44.005}^{\:70.295}$ & $58.230_{\:54.026}^{\:62.826}$ \\
		& $\tau_{2}$ & $8.909$ & $10.445_{\:7.396}^{\:15.490}$ & $9.353_{\:8.387}^{\:10.399}$ & $8.970_{\:8.681}^{\:9.287}$ \\
		\hline
		\hline
		State force mean (pN) 		& $\mu_{1}$ & $3.000$ & $4.677_{\:4.644}^{\:4.712}$ & $5.606_{\:5.600}^{\:5.613}$ & $3.002_{\:2.991}^{\:3.013}$ \\
		& $\mu_{2}$ & $4.700$ & $3.069_{\:2.953}^{\:3.187}$ & $3.019_{\:2.986}^{\:3.056}$ & $5.602_{\:5.600}^{\:5.604}$ \\
		& $\mu_{3}$ & $5.600$ & $5.602_{\:5.580}^{\:5.625}$ & $4.700_{\:4.690}^{\:4.710}$ & $4.703_{\:4.699}^{\:4.706}$ \\
		\hline
		State force std dev (pN) 		& $s_{1}$ & $1.000$ & $0.311_{\:0.285}^{\:0.337}$ & $0.199_{\:0.194}^{\:0.203}$ & $1.002_{\:0.995}^{\:1.010}$ \\
		& $s_{2}$ & $0.300$ & $1.067_{\:0.992}^{\:1.147}$ & $1.012_{\:0.988}^{\:1.039}$ & $0.200_{\:0.199}^{\:0.202}$ \\
		& $s_{3}$ & $0.200$ & $0.210_{\:0.195}^{\:0.229}$ & $0.297_{\:0.290}^{\:0.305}$ & $0.299_{\:0.296}^{\:0.301}$ \\
		\hline
\hline
\end{tabular*}
\end{table*}
